\chapter{Philosophy's Copernican Revolution: Time in The \textit{Critique of Pure Reason}}

\chapterprecishere{The history of Immanuel Kant's life is difficult to portray, for he had neither life nor history ... I do not believe that the great clock of the cathedral performed in a more passionless and methodical manner its daily routine than did its townsman, Immanuel Kant. Rising in the morning, coffee drinking, writing, reading lectures, dining, walking, everything had its appointed time, and the neighbours knew that it was exactly half-past three-o-clock when Immanuel Kant stepped forth from his house in his grey, tight fitting coat, with his Spanish cane in his hand, and betook himself to the little linden avenue called after him to this day the `Philosophers walk. "... What a strange contrast did this man's outward life present to his destructive, world annihilating thought! In sooth, had the citizens of Konigsberg had the least presentiment of the full significance of his ideas, they would have felt a far more awful dread at the presence of this man than at the sight of the executioner, who can but kill the body. But the worthy folks saw in him nothing more than a Professor of Philosophy, and as he passed in his customary hour, they greeted him in a friendly manner and set their watches by him\par\raggedleft--- \textup{Heinrich Heine}, Religion and Philosophy in Germany, \textup{p. 109.}}

\vskip 5ex
% TODO: Fix quote/epigraph format.
\chapterprecishere{\par\raggedleft\textbf{\textup{49. Ko. Revolution.}} \begin{flushleft}\textbf{\textup{[Molting]}}\end{flushleft}\ Fire in the water. The image of REVOLUTION. Thus the superior man sets the calendar in order and makes the seasons clear.\par\raggedleft---I Ching: The Book of Changes, \textup{p. 190.}}

\section{Introduction: The Discovery of Transcendental Time}

In 1781 Immanuel Kant published the first edition of the \textit{Critique of Pure Reason}. The history of philosophy registers the date as the moment of Kant's "Copernican revolution, " a moment in which Kant is said to have accomplished in the realm of thought what Copernicus had accomplished, over two centuries earlier, in the realm of astronomy. Traditionally, the story of Kant's revolution emphasizes epistemology and concentrates on the role of the human intellect in constituting the external world. It is here, we are told, that one should locate the dramatic shift that is at the core of Kantian philosophy.\footnote{ In the following section this position will be illustrated through the writings of Heinrich
Heine. Though Heine poetic and lyrical language is unique amongst the
commentators on Kant, his views are not at all unconventional. In the '"Past Masters" text on Kant, for example, Roger Scruton, writes that the essence of the Copernican revolution is that 'self consciousness requires that the world must appear to conform to the categories' (Scruton, Roger. \textit{Kant}. Oxford: Oxford University Press, 1982, p. 28). The Encyclopedia of Philosophy concurs that this epistemological view is the key to understanding Kantian thought. 'Kant's principle task in the \textit{Critique of Pure Reason} was to determine the cognitive powers of reason, to find out what it could and could not achieve in the way of knowledge" ("Kant, Immanuel" in the \textit{Encyclopedia of Philosophy Volume 3}. Edited by Paul Edwards. New York: Macmillan Publishing, 1967, p. 305).} What this chapter will argue, however, is that this emphasis, on the role of the intellect, mistakes what is truly revolutionary in critical thought. The argument, which draws on Deleuze's reading of Kant,\footnote{This is comprised both of Deleuze's book on Kant and a series of lectures on Kant that have been reprinted on the Internet (see bibliography).} consists of two main points. First, that the transformation of the human subject is merely a consequence of Kant's more fundamental innovation, the discovery of the realm of the transcendental, and second, that to appreciate the truly revolutionary nature of this discovery we must turn our attention away from the enlightened subject of reason and focus instead on the occulted nature of time.

The most common approach to Kant is to read him as an epistemologist, a philosopher who is predominantly concerned with how knowledge can be justified. "According to this account \textit{The Critique of Pure Reason} centers around the question: "How are \textit{a priori} synthetic judgements possible?" (CPR, 59). To answer this question Kant must begin by defining what he means by the synthetic \textit{a priori}. 

"In the order of time," writes Kant in the preface to the first Critique, "we have no knowledge antecedent to experience, and with experience all our knowledge begins. But though all our knowledge begins with experience it does not follow that it all arises out of experience" (CPR, 41). With this distinction---between knowledge that is based in experience and knowledge that is independent of it---Kant separates the \textit{a priori} from the \textit{a posteriori}. This difference is absolute and rigorously determined. Empirical knowledge which has its sources \textit{a posteriori}, that is, in experience" (CPR-43) gives rise to judgments which are particular and contingent (the sun rose today) while \textit{a priori} knowledge, on the other hand, gives rise to judgments which are universal and necessary (today succeeded yesterday). Since knowledge gained from experience is always particular and contingent it can never be the basis for judgments which are universal and necessary. Thus, Kant writes, "necessity and strict universality are the sure criteria of \textit{a priori} knowledge" (CPR, 44). This difference between the \textit{a priori} and the \textit{a posteriori} is the first basic division which allows Kant to demarcate the singular zone of knowledge that he is concerned with in the first Critique.

Yet this distinction, between knowledge that is based in experience and knowledge that is independent of it, is not in itself sufficient for understanding the central problematic of Kantian thought. In order to discover the transcendental, yet another distinction was required. The two sides of the table had to be split in half. To accomplish this, Kant drew another line on a different axes, cutting across both the \textit{a priori} and the \textit{a posteriori}.\footnote{See Table 1 in the appendix.} This line corresponds to the difference between knowledge that is analytic and knowledge that is synthetic. 

Analytic \textit{a priori} cover logical truths. By drawing out "something that is (covertly) contained within a concept" (CPR, 48) they analyze, elucidate, or explicate what is already implicitly known. Kant's example is as follows: "If I say, for instance 'All bodies are extended' this is an analytic judgment. For do not require to go beyond the concept which I connect with 'body' in order to find extension as bound up with it" (CPR, 48).

Analytic knowledge is restricted to the domain of the a priori. It does not arise empirically. For "judgments of experience, " writes Kant, "are one and all synthetic" (CPR, 49) By the term 'synthetic' Kant is referring to knowledge that 'goes beyond the concept'. To make the judgment 'this body is heavy', for example, is to connect concepts 'synthetically'. For since not all bodies are heavy, one cannot arrive at the concept heavy from an analyses of the concept 'body'.\footnote{On page 49 of the first critique Kant uses this example to explain synthetic judgements which are made from experience.}


% Fix Math handling.
Transcendental philosophy concerns itself with the final box in the table---the one that contains the 'synthetic \textit{a priori}'. The puzzle which it sets for itself is how synthetic knowledge can be produced independently of experience. The prime examples of synthetic a priori knowledge are found within the realm of mathematics. As the commentator Alfredo Ferrarin writes, "If what the Critique shows is the possibility of synthetic \textit{a priori} judgements, it is mathematics that takes advantage of this ampliative principle with greatest confidence and success" (KECI, 147.) '7+5=12', to stick with the example that is repeated throughout the secondary literature on Kant,\footnote{Kant uses this example himself both in the first critique and in the \textit{Prolegomena to Future Metaphysics}.} is a synthetic \textit{a priori} judgment. It is \textit{a priori} since it, like all mathematical propositions, is both universal and necessary. It is synthetic since neither the number 7 nor the number 5 has contained within it the number 12. "The concept of 12," writes Kant, "is by no means already thought in merely thinking the union of 7 and 5; and I may analyze my concept of such a possible sum as long as I please, still I shall never find the 12 in it" (CPR, 53).

Though philosophers had long been concerned with such a priori truths as are found in mathematics it was Kant who first recognized them as being synthetic.\footnote{As Deleuze writes, "analytic a priori judgment that meant something, synthetic a posteriori judgment that meant something, but synthetic a priori judgment, that's truly a monster" (KST, 9).} Thus transcendental philosophy, even when confined to epistemology, is far from being a mere exercise in the catalogue of knowledge. For in questioning the tendency to divide everything between the \textit{a priori} analytic judgment and the \textit{a posteriori} synthetic judgment Kant "exposed the insufficiency of certain philosophical categories and conceptual frames" (KST, 9). Focusing his attention on the synthetic \textit{a priori} he exploded old distinctions and in so doing discovered a powerful new machine. 

According to Kant, synthesis is "a blind but indispensable function of the soul". Though "we are scarcely ever conscious" of its power, without it, he insists, "we should have no knowledge whatsoever" (CPR 112). In focusing on the connections and constructions of this hidden realm, the \textit{Critique of Pure Reason} develops a 'synthesized way of handling philosophy'\footnote{This phrase comes from a question raised by Deleuze in an introductory lecture on Kant. "Why," Deleuze asks, "wouldn't there also be a synthesized or electronic way of handling philosophy" (SLK, 1).} which is not based on analyses of that which is already given but on an "extension of our previously possessed concepts" (CPR, 47). It is, writes Kant "a genuinely new addition to all previous knowledge?" (CPR, 51).

It is this 'new addition' that accounts for the shift in the subject's position which occurs in Kantian thought. What is important to recognize here, is that this shift is a result, a corollary, of the more fundamental discovery of this abstract and productive realm of knowledge. Before Kant, the subject was found buried, submerged underground, chained in the darkness of Plato's cave. According to this traditional vision, the subject was trapped in the body, forced to access the world through the unreliability of the perceptual apparatus. An unfortunate fool, blinded by ignorance, duped into mistaking shadows for reality, the subject could not help but deform the world, mutating it into the falsity of illusion. Philosophy's striving consisted in its promise to provide the escape route. Operating with a truth that depended on a "harmony between the subject and the world," (KST, 5) philosophy struggled to cut the chains, to correct the inherent deformity, to free the prisoner from the world of shadows and illusions. 

Through his discovery of the transcendental, Kant replaces harmony with circuitry. The subject, no longer deceived and defective, becomes productive and constitutive. Having given up the impossible attempt at conforming to the objects of the world, "the rational being discovers he has new power" (KCP, 14). After Kant the objects of the world must conform to us. "The first thing the Copernican Revolution teaches us," writes Deleuze, "is that it is we who are giving the orders" (KCP, 14). The prisoner has become a legislator. 

As previously noted, it is this shift in the subject's position which is traditionally taken to be at the core of Kant's revolutionary thought. In a dramatic passage, the poet Heinrich Heine, describes this, philosophy's Copernican revolution, as follows:

\begin{quote}
    Formerly, when men conceived the world as standing still, and the sun as revolving round it, astronomical calculations failed to agree accurately. But when Copernicus made the sun stand still and the earth revolve round it, behold! Everything accorded admirably. So formerly reason, like the sun moved round the universe of phenomena and sought to throw light upon it. But Kant, bade reason, the sun stand still, and the universe of phenomena now turns round, and is illuminated the moment it comes within the region of the intellectual orb (RPG, 114).
\end{quote}

Yet, while Heine has captured the drama of Kant's discovery, his account reveals a certain problem. For if the stress is on human reason, Kant's allusions to Copernicus are somewhat puzzling. Before Copernicus, Heine reminds us, the earth stood as the central pivot or axis around which everything else revolved. Modem astronomy, which is based on the Copernican system, removed the earth from this central position, making it equal to any other planet. The Copernican revolution thus derailed us from our privileged status in relation to phenomena. Kant, on the other hand, is said to have done the exact opposite. Whereas Copernicus displaced us from the center of the universe, Kant put us there. Why, then, does Kant speak of his philosophy as Copernican? For it would seem that the emphasis on the human intellect is not a sufficient explanation. Perhaps, if we look more closely, we might find some other reason for this seemingly confused analogy. 

In the early years of the 16th century, Nicolaus Copernicus, a polish astronomer, attained immortal fame by overthrowing the Ptolemaic universe of the ancient world. Frustrated with the impossibility of achieving accurate measurements of astronomical movements, Copernicus began to question Ptolemy's geocentric vision. Instead of assuming that the stars revolved around a still earth, Copernicus thought, to quote Kant, "whether he might have better success if he made the spectator to revolve and the stars to remain at rest" (CPR, 23). Copernicus posited a heliocentric world in which the stars no longer measured time. He explained the day by the earth's rotation on its own axes and the year by its annual cycle around the sun. In the Copernican system, then, it is the movement of the earth which marks out the temporality of the astronomical calendar. 

The \textit{Critique of Pure Reason}, writes Kant, proceeds "precisely on the lines of Copernicus' primary hypothesis" (CPR, 23). Inspired by the astronomers' method, Kant attempted an analogous experiment in philosophy. Frustrated by the inherent instability of metaphysics, critical thought seeks to attain more solid foundations by focusing not on the authority of experience but on the conditions which make experience possible. 

It is well known that Copernicus' discovery met with fierce resistance, both from natural philosophy and from the church. For these two institutions were allied in their commitment to maintaining the authority of Aristotle, who had insisted that the earth stood still. This resistance was heightened by the fact that despite Copernicus' findings the world still appeared to conform to Ptolemy's ancient vision. Copernicus was thus responsible for a strange and mysterious revolution in which nothing seemed to change but through which everything has been transformed. It is in this way, as we will see, that Kant is a true Copernican. For the Copernican revolution, whether in astronomy or philosophy, changes nothing at the level of experience. Our perceptions and even the way we talk about those perceptions have not altered. Phenomena remain the same. The sun still appears to revolve. The earth still appears to stand still. External bodies still appear to be in motion. We still say that the sun rises and sets. The difference is, and this is the revolution, that now everybody knows it is only a manner of speaking.

In a series of lectures on the \textit{Critique of Pure Reason}, Gilles Deleuze maps out a singular and original account of Kant's Copernican revolution which is based neither on epistemology nor on a change in the position of the intellect but on a shift in the nature of appearance itself. 

According to Deleuze, the classical tradition structured the world around a basic opposition. "The whole of classical philosophy from Plato onwards," he writes, "seemed to develop itself within the, frame of a duality between sensible appearances and intelligible essences" (KST, 4). Thus, previous to Kant the world was divided between, on the one hand, the degraded realm of sensation which was based on bodily knowledge and experience, and on the other hand, the realm of ideas, pure forms or essences which were transcendent and therefore untainted by the blemishes of sensation. 

For Plato \textit{a priori} knowledge was the proof of transcendence. His dialogues insist that the very fact that there is knowledge independent of experience shows that reason remembers a time when it was unfettered by the body's cage and was free to gaze upon the pure essence of things.\footnote{One of the most famous examples of the Platonic view of the a priori occurs in the \textit{Meno} where Socrates infers the transcendence of the Forms through a slave's knowledge of geometry. See Plato, "Meno" in \textit{Plato---The Collected Dialogues}. edited by Edith Hamilton and Huntington Cairns. Princeton New Jersey: Princeton University Press, 1961 (especially pages 363-374).} The philosophical distinction between \textit{a posteriori} and \textit{a priori} knowledge was thus, for Plato, evidence of the fact that our capture in the illusory realm of phenomena could be opposed to an exteriority characterized by the transcendent truth of the idea.

The \textit{Critique of Pure Reason} overturns the classical tradition by developing a philosophy that is no longer grounded in this basic opposition. "For the disjunctive couple appearance/essence," writes Deleuze, "Kant will substitute the conjunctive couple what appears/conditions of appearance. Everything is new in this" (KST 4).

With Kant, then, phenomena cease to be trapped by the ancient duality. "It's like a bolt of lightening" (KST, 4). The world of appearances vanish. What is left instead, according to Deleuze, is the apparition. "The apparition, is what appears in so far as it appears. Full stop. I don't ask myself if there is something behind, I don't ask myself if it is false or not false. The apparition is not at all captured in the oppositional couple, in the binary distinction where we find a appearances distinct from essences" (KST, 4). 

No longer bound by the fundamental distinction of classical thought, Kant transforms the meaning and implications of \textit{a priori} knowledge. "In the case of the \textit{a priori}, " writes Deleuze, "Kant "borrows a word but he completely renews its sense. " (KST, 1) For unlike Plato, in Kant the \textit{a priori}, as we will see, is associated with the immanence of abstraction and not the transcendence of the eternal forms. To quote from Deleuze: "Kant is the one who discovers the prodigious domain of the transcendental. He is the analogue of the great explorer - not of another world, but of the upper and lower reaches of this one" (DR, 135). In opposition to the transcendent ideas and logic of the analytic \textit{a priori}, the synthetic \textit{a priori} constitute a continuous process of production that is both exterior and immanent to our experience of the world. The basic question of transcendental philosophy "how are \textit{a priori} synthetic judgments possible?" can thus be restated as follows: given a certain experience what are the conditions that went in to producing it? Kant's answer, as we will see, shows that that which is exterior---or independent---of experience is not a transcendent world above us but rather an immanent outside. It is this which he calls the transcendental.\footnote{According to Deleuze, "The whole Kantian notion of the transcendental is created in order to refute the classical notion of the transcendent. The transcendental is above all not the transcendent" (SLK, 7). }

As was noted in the introduction, the classical disjunction between essence and appearance corresponds to the distinction in the philosophy of time between, on the one hand, the phenomena of temporality and change, and, on the other, the essence of eternity. Overturning this classical duality between essence and appearance requires not only a transformation in way we approach phenomena but also a fundamental reinvention in the philosophy of time. As we will see in the chapter which follows, the discovery of the domain of the transcendental ultimately rests on this reinvention---or revolution---in the nature of time. Deleuze, recognizing this, writes in an introduction to Kant that "all the creations and novelties that Kantianism will bring to philosophy turn on a certain problem of time and an entirely new conception of time (KST, 1). It is to this new conception of time which this chapter now turns.
\newpage

\section{Transcendental Aesthetic: Time as the Form of Inner Sense}
\vskip 4ex
\chapterprecishere{Until now the task we have given ourselves was to represent space, the
moment has come to think time.\par\raggedleft--- \textup{Gilles Deleuze}, Kant: Synthesis and Time, \textup{p. 1.}}

With its central divisions, parts, sections, chapters, books, sections of chapters, and chapters of books, the structure of the \textit{Critique of Pure Reason} seems more like the work of a ramshackle artificial intelligence than that of a human being.\footnote{De Quincy's text 'The Last Days of Immanuel Kant' gives further evidence of this seemingly preposterous claim. Besides the meticulous order of his daily schedule, Kant never perspired, evoked rigorous numerological arrangements for the guests at his dinner table, and, at his deathbed, when all human faculties had left him, was still able to speak at length on any problem in history, philosophy or mathematics. See: De Quincy, Thomas. 'The Last Days of Immanuel Kant' in \textit{The Collected Writings of Thomas De Quincy} Vol. IV. London: A\&C Black, 1897.} The immense scale and complexity of Kant's 'thinking machine' is revealed with one glance at the table of contents. 

The bulk of the text is divided into two main parts, the 'Transcendental Aesthetic' and the 'Transcendental Logic'. This split corresponds to the central distinction in Kantian thought which divides the intuition from the understanding. Intuition deals with the realm of sensation. Receptive and immediate, it is the form in which the diversity of sense material is presented to the mind. Understanding, on the other hand, is defined as the 'spontaneous production of concepts' (CPR, 92). An active mediation, rather than a receptive and immediate presentation, it serves to represent the perceptions that are given to us in intuition in accordance with the categories of reason. Time appears first in the former category. It is defined in the opening section of the \textit{Critique of Pure Reason} as a 'pure intuition,' or the 'form of inner sense' (CPR, 77).

The Transcendental Aesthetic, "the science of a priori sensibility"
(CPR, 66) begins with a strict process of elimination. It is concerned only with what is left after both the concepts of the understanding and the matter of sensibility have been stripped away. What remains are what Kant calls the 'pure intuitions,' or 'the form of appearances'. These are defined as the underlying conditions that constitute our perception of the world. The transcendental media for the reception of sensible content, they constitute the structure and form which the apparition must take. For Kant there are only two such forms, space and time. 

Kant defines space as "the form of outer sense". It is "the property of our mind" through which "we represent to ourselves objects which are outside us" (CPR, 67). Thus, for Kant, space is the form in which the external world is presented to the senses. That is to say, everything that we sense as external to us we necessarily perceive of as being in space. To quote from
the first Critique: 

\begin{quote}
    In order that certain sensations be referred to something outside me (that is, to something in another region of space from that in which I find myself), and similarly in order that I may be able to represent them as outside and alongside one another, and accordingly as not only different but as in different places, the representation of space must be presupposed (CPR 68).
\end{quote}

Time, on the other hand, is defined as "the form of inner sense, that is of the intuition of ourselves and of our inner state" (CPR, 77). In the 'Transcendental Aesthetic,' then, time provides the underlying structure of all our states of mind. It is what conditions the very experience of thought, including our awareness of outer perceptions and the consciousness we have of ourselves. As Kant writes, "everything which belongs to inner determinations is represented in relation to time" (CPR, 68).

In the following section we will see that in making time the form of inner sense Kant revolutionizes the classical philosophy of time both by liberating it from its dependence on change and by releasing it from an implicit spatial bias. 

According to Deleuze, the first great Kantian reversal in the \textit{Critique of Pure Reason} was to free time from its subordination to movement Taking Hamlet's phrase the 'time is out of joint' and applying it to Kant,\footnote{In his text "On four poetic formulas that might summarize the Kantian philosophy" (found in both \textit{Kant's Critical Philosophy} and in \textit{Essays Critical and Clinical}) Deleuze uses this Shakespearean quote---'the time is out of joint'---to explore Kantian thought. To quote Deleuze: "Hamlet is the first hero who truly needed time to act, whereas earlier heroes were subject to time as the consequence of an original movement (Aeschylus) or an aberrant action (Sophocles). The Critique of Pure Reason is the book of Hamlet, the prince of the north" (ECC, 28).} Deleuze shows how in taking time off its hinges, Kant develops a "sort of modern consciousness of time" (KST, 1). In this 'modern consciousness' time is separated from the external world of space and thus undergoes a sort of  topological twist. What was once located in the external world is folded in.\footnote{To quote Kant: "Time is not something which exists of itself, or which inheres in things as an objective determination, and it does not, therefore, remain when abstraction is made of all subjective conditions of its intuition\dots Time is nothing but the subjective condition under which alone intuition can take place in us" (CPR, 76).} Time, detached from the movement of that which is outside us becomes the structuring principle which conditions the inside. Thus, as we will see, in making time the form of inner sense Kant not only redefines the classical conception of time, he also transforms the traditional understanding of interiority and its relation to the outside. 

In Plato's dichotomized world, time exists on only one side of the mirror. Essences, which are eternal and real, exist in a transcendent realm outside time. Time, on the other hand, belongs to the world of appearances, which, according to Plato, is governed by a continuous process of change and movement. For Plato, then, it is change and movement that are the defining features of time. 

In his dialogue Timeaus, Plato describes time as the "image of eternity" (T, 1167). He perceives this image in the movement of the stars and thus equates the production of time with the 'perfect and immutable' cycles within which the planets revolve. "Such was the mind and thought of God in the creation of time. The sun and the moon and five other stars which have the name of planets were created by him in order to distinguish and preserve the numbers of time" (T, 1167). In the 'curved time' of a 'circular universe' Plato's God has bent the skies into an arc. "A Demiurge which makes  circles," (SLK, 2) as Deleuze puts it, has created a world whose map is observed in the heavenly spheres. Thus, for Plato, 'the name' and 'the number' of time can be found in the changes and motion that take place in the sky. 

Kant's break with Plato is absolute. For, according to the first Critique, movement is empirical, which is to say it exists at the level of experience. Transcendental philosophy, he writes, "cannot count the concept of alteration among its \textit{a priori} data" (CPR, 75). For Kant the revolutions of the stars, the swing of a pendulum, the sand in an hourglass all occur \textit{in} time and as such fall outside the problematic of critique. Plato's perfect image can not be time, for, "time itself does not alter but only something which is in time" (CPR, 82). According to Kant, time explains the possibility of movement, but movement is not time. "The concept of alteration and with it the concept of motion, as alteration of place," he writes, "is possible only through and in the representation of time" (CPR, 75).

Thus, Kant liberates the form of time from the endless cycle of the Platonic world. Time is no longer contained within the circular revolutions of the planets. Instead, it is the relation of time to motion itself that revolves. In Kant, writes Deleuze, "time is no longer related to the movement which it measures, but movement is related to the time which conditions it". The \textit{Critique of Pure Reason} conceives of a 'straight line' of time that is cut off from its subordination to all that exists in time.\footnote{It is interesting to note that Foucault, when writing of Deleuze, evokes the 'straightening' of time that is initially produced by the first Critique. The circle must be abandoned as a faulty principle of return; we must abandon our tendency to organize everything into a sphere. All things return on the straight and narrow by way of a straight and labyrinth line" (TP, 166). This notion of the straight labyrinth of time is used by Deleuze to designate the transcendental form of time. In the preface to \textit{Kant's Critical Philosophy}, he writes as follows: We move from one labyrinth to another. The labyrinth is no longer a circle, or a spiral which would translate its complications, but a thread, a straight line, all the more mysterious for being simple, inexorable as Borges says, "the labyrinth which is composed of a single straight line, and which is indivisible, incessant" (KCP, vii). Deleuze also uses the notion of a straight labyrinth to describe the time of Aeon (a concept that will be discussed in detail in chapter 3).} To quote from Deleuze:

\begin{quote}
    Time is no longer coiled up in such a way that it is subordinated to the measure of something other than itself, such as, for example, astronomical movement. Everything happens as if, having been coiled up so as to measure the passage of celestial bodies, time unrolls itself like a sort of serpent, it shakes off all subordination to a movement or a nature, it becomes time in itself for itself, it becomes pure and empty time. It measures nothing anymore. Time has taken on its own excessiveness. It is out of its joints, which is to say its subordination to nature; its now nature which is subordinated to it (KST, 12).
\end{quote}

Since time is already presupposed in motion, Kant, 'the great explorer,' must begin to search elsewhere. Looking behind objects to discover the conditions of their production, Kant finds that time is "not an empirical concept that has been derived from any experience" (CPR, 74). In the abstract realm of the transcendental, Kant discovers a form of time that is independent of the experience of motion.

It would first appear that Kant finds a predecessor in Aristotle who modifies the Platonic vision, by making a distinction between motion and time. Though Aristotle connects time to the cycles of astronomical change he does not equate it with them. His argument rests on two points. First, that while time exists everywhere, movement only occurs in particular things. Second, that while things that move can be either fast or slow, time itself does not shift in speed. "It is evident, then," writes Aristotle, "that time is neither movement nor independent of movement" (P, 371).

In order to discover the precise relation between motion and time, Aristotle turns to number. To quote his famous formula "time is just this---the
number of motion" (P, 372). The circular revolutions, the heavenly spheres are still linked to time but the two are no longer directly equivalent. "Time, is only movement," writes Aristotle "in so far as it admits of enumeration" (P, 372). Change must be numbered for it to be time. 

"Time, then," for Aristotle, "is a kind of number" (P, 372). The kind of number that it is, rests on a distinction made in the Physics, between the "number with which we count", and "the number of things which are counted" (P, 373). To quote from the \textit{Physics}: "Number, is used in two ways---both of what is counted or countable and also of that which we count. Time, then, is what is counted, not that with which we count: these are different kinds of thing (P, 372).

Thus, Aristotle, like Kant, seeks to discover time by shifting focus away from the concreteness of astronomy. However, as we will see, the abstraction he makes in the direction of number is ultimately subordinated to empirical movement. For according to Aristotle the numbers of time are determined as "the measure of a quantity of change" (P, 375).

For Kant, as for Aristotle, time is conceived as being fundamentally numeric. Kant recognizes that any representation cannot help but serve to spatialize time. Since time 'yields no shape' (CPR, 77) it cannot be represented. Still, he focuses on a single spatial image that will function as an analogy for his new form of time. For Kant, the closest we get to time in space, is the image of the real number line.\footnote{Commentator Alfredo Ferrarin makes much of this, arguing that in Kant numbering and arithmetic are practically synonymous with the generation of time.} "We represent the time sequence," he writes, "by a line progressing to infinity, in which the manifold constitutes a series of one dimension only; and we reason from the properties of this line to all the properties of time" (CPR 77).\footnote{This sentence continues as follows: "with one exception that while the parts of the line are simultaneous the parts of time are always successive" (CPR, 77). Yet, it is important to note, that while the parts of time are successive, time itself, for Kant, is not. Deleuze is insistent on this point. "Time," he writes, "is no longer defined by succession because succession concerns only things and movements which are in time. If time itself were succession, it would need to succeed in another time, and on to infinity. Things succeed each other in various times, but they are also simultaneous in the same time, and they remain in an indefinite time. It is no longer a question of defining time by succession, nor space by simultaneity, nor permanence by eternity. Permanence, succession and simultaneity are modes and relationships of time" (KCP, vii-iii).}

Yet, despite the fact that both Kant and Aristotle link time to number, Kant breaks with Aristotle no less than with Plato. For in freeing time from its subordination to movement, Kant had to uncouple number from measurement. In the domain of the transcendental, the numerical processes of time are unhinged from the concreteness of change. With Kant the number of time thus breaks from Aristotle's classical formula. Ceasing to function as 'the quantity of motion', the 'things that are counted' become 'that with
which we count'. 

It is easier to understand what is at stake in this reversal if we map Aristotle's distinction in the philosophy of number on to the difference between cardinal and ordinal numbers. Cardinal numbers [one, two, three\dots] are used to express amount or quantity. Ordinal numbers, on the other hand, are used to express position [first, second, third]. Bound to keep track of that which it belongs to, the cardinal number is tied down, attached to what is in time. "Cardinal," writes Deleuze, "comes from cardo; cardo is precisely the hinge, the hinge around which the sphere of celestial bodies turns and which makes them pass time and again through the so-called cardinal points" (KST, 12). Ordinal numbers, on the other hand, are indifferent to the space of measurement.\footnote{A familiar example of ordinal numbering is that used by the library cataloguing system. It is clear, in this case, that what is important is the order of the books and not the amount of space between them.} What counts, for them, is order not measure. In unchaining time from its bonds to what is in time, Kant simultaneously freed number from its subordination to measurement. As Deleuze writes, in Kant's 'new definition of time' number "ceases to be cardinal and becomes ordinal, a pure order of time" (DR, 88).

In the end therefore, despite the differences between Plato's and Aristotle's conceptions of time, from the perspective of transcendental philosophy, they are basically alike. For both belong to "a certain tradition of antiquity, in which time is fundamentally subordinated to something which happens in it, and this something can be determined as being change" (SLK, 1). By separating time from the heavenly spheres and making number independent of motion, Kant thus splits with an entire tradition. Operating with the conjunction 'what appears/conditions of appearance,' Kant creates (or discovers) a disjunction between the form of time itself and the changes which occur in time.

This distinction between the abstract form of time and the changes which occur within it requires that time be released from its spatial determinations. In the \textit{Critique of Pure Reason}, time is no longer located in the objects of the world or in their relations, but is situated instead in an abstract realm that is independent from our perceptions of space. The transcendental form of time, as we have seen, is an empty form, conceived of as nothing but a pure ordinal sequence. Since time has no spatial dimension, even the number line is but an analogy.

In the Transcendental Aesthetic, space, as we have seen, captures the whole of exteriority inside itself. Thus in order to separate time from space, time which was once located in the external world, must be folded in. For, according to the first Critique, the only thing that is not in space is the inner determinations of our mind. In making time the 'form of inner sense' Kant thus locates time in the singular domain which exists outside the representation of space.

This process of interiorization gives time a certain dominance over space. For according to Kant everything that we represent as in space must also be processed by time, precisely insofar as it is experienced (and thus belongs to our inner states). "Appearances," writes Kant, "may, one and all, vanish; but time (as the universal condition of their possibility) cannot itself be removed" (CPR, 75). Thus, for Kant, while everything that exists is in time, time is the one thing that does not exist in space, since everything in space 'already' presupposes time.\footnote{From the perspective of transcendental philosophy, then, William Burroughs' formula, "The only way out of time is into space" (H, 19) can only be a trick.} To quote from the first Critique:

\begin{quote}
    Time is the formal \textit{a priori} condition of all appearances whatsoever. Space, as the pure form of all \textit{outer} intuition, is so far limited; it serves as the \textit{a priori} condition only of outer appearances. But since all representations, whether they have for their objects outer things or not, belong in themselves, as determinations of the mind, to our inner state; and since this inner state stands under the formal condition of inner intuition, and so belongs to time, time is an \textit{a priori} condition of all appearances whatsoever\dots Just as I can say \textit{a priori} that all outer appearances are in space, and are determined \textit{a priori} in conformity with the relations of space, I can also say, from the principle of inner sense, that all appearances whatsoever, that is, all objects of the senses, are in time, and necessarily stand in time relations (CPR, 77).
\end{quote}

Once abstracted and interiorized time takes on enormous new powers. Productive of the actual rhythm of thought and sensation it gains control over the whole of experience. Time, as the form of interiority, is thus absolutely inescapable. For everything we see, think, feel, hear and know has already been given a speed, an order and a rhythm in time. With Kant, then, the inside ceases to be conceived of an empirical container and is instead thought transcendentally (as an interiority over against space and not merely in space). One can never escape time, since time is a limit that works us from the inside. Yet, as Deleuze notes, there is something very strange in this notion of time as interior limit.

\begin{quote}
    To think time means to substitute for the classical schema of an exterior limitation of thought by the extended, the very very strange idea of an interior limit to thought which works it from the inside, which doesn't at all come from outside, which doesn't at all come from the opacity of a substance. As if there was in thought something impossible to think. As if thought was worked over from the inside by something that it cannot think (TLK, 1).
\end{quote}

In moving from transcendence to the transcendental, Kant reworks both the inside and the outside. Time conditions an inescapable interiority, but in doing so opens a new and more radical exteriority, since the production of time itself cannot be captured "within" time. In other words, the one thing that is not interior to time is the transcendental form of time itself. Thus, in discovering the abstract realm of the transcendental, Kant unmasks an unanticipated immanent exteriority---an outside that does not transcend the world but that is no less alien for that. "The greatest initiative of transcendental philosophy," writes Deleuze, "was to introduce the form of time into thought" (DR, 86). Yet, as the next section will show, it is only a very particular mode of thought that can process Kant's modern consciousness of time. For Kant revolutionizes the interiority of thought only by immersing it in the exteriority of time.

\newpage
\section{Transcendental Deduction: Time and the `I think'}

\vskip 4ex
\chapterprecishere{How can man think what he does not think, inhabit by a mute occupation
something that eludes him, animate with a kind of frozen movement that
figure of himself that takes the form of a stubborn exteriority.\par\raggedleft--- \textup{Michel Foucault}, The Order of Things, \textup{p. 323}}

\vskip 4ex
\chapterprecishere{It is false to say: I think. One should say: one thinks me\dots I is another\par\raggedleft--- \textup{Arthur Rimbaud}, Complete Works, \textup{p. 101}}

\vskip 4ex
\chapterprecishere{I am separated from myself by the form of time.\par\raggedleft--- \textup{Gilles Deleuze}, Kant's Critical Philosophy, \textup{p. ix}}


In modern philosophical thought it is Descartes who first secures the borders which separates the inside from the outside. His book, the \textit{ Meditations}, assures the divide by devising what has become one of philosophy's most familiar horror stories.\footnote{As Slavoj Žižek writes, the Cartesian cogito "opens up, for a brief moment, the hypothesis of the Evil Genius who, behind my back, dominates me and pulls the strings of what I experience as reality". Slavoj Žižek. \textit{Tarrying with the Negative: Kant, Hegel and the Critique of Ideology}. Durham: Duke University Press, 1993, p. 12} Deeply suspicious of appearances, Descartes speculates that God has been replaced by an evil demon who tricks us into taking the exterior world for reality. Determined to escape these demonic delusions, Descartes supposes "that the heavens, the air, the earth, colours, shapes, sounds and all external things that we see, are only illusions and deceptions which he [the demon] uses to take me in" (M, 100). Stripped of all faith in reason and truth, Descartes imagines the external world as a mad and uncontrollable experiment. The senses, which grant us access to this outside world, must no longer be trusted. Even the certainties of abstract truths such as are found in mathematics and geometry must be questioned. For once the demon has been let in, there is no telling how far his influence extends. Plagued by a powerful and cunning entity bent on deception, Descartes tries to flee from the demon, demarcating the outside by a rigorous process of philosophical doubt.

Yet, as time passes, Descartes manages to gain control and dispel the horror which torments him. Turning in on himself, Descartes discovers that the doubt which haunts the outside serves to ground the inside in an undeniable certainty.\footnote{To quote Descartes: "There is a deceiver---I know not who he is---most highly powerful and most highly cunning, who always industriously deceives me. If he is deceiving me then without doubt I also am. And he might deceive me as much as he can, he will still never effect that I would be nothing, so long as I shall cogitate that I am something. So that---all things having been weighed enough, and more---this statement were, finally, to be established: "\textit{I am, I exist}" is necessarily true, so often as it is uttered by me or conceived by the mind"(Meditation 11: 3).} The fact that I doubt, the \textit{Meditations} famously concludes, assures me that I exist. For, "when I doubt, there is one thing which I cannot doubt, which is that as a self that doubts, I think" (SLK, 2). In doubting the objects, properties, truths and sensations of the external world, Descartes thus carves out a secure container for the subjects insides. The famous formula "I think therefore I am" guards us against the demonic games that threaten our perception and knowledge of the outside world.

Separated from the dubious nature of exteriority, the Cartesian subject thus finds its protection by folding in. For though everything outside us may be a trick, nothing, Descartes contends, can fool us about what exists inside. Guarded by the security of self consciousness, Descartes uses the boundaries of a conscious interiority to rebuild his faith in the external world. After Descartes, it is this certainty of interiority which has became the sure foundations for philosophical inquiry.\footnote{As Deleuze writes, in Descartes "doubt was only a deter to a more secure edifice of knowledge" (DD, 77).} Schopenhauer describes this as follows:

\begin{quote}
    By his taking cogito ergo sum as the only thing certain, and provisionally regarding the existence of the world as problematical, the essential and only correct starting point, and at the same time the true point of support, of all philosophy was really found. This point, indeed, is essentially and of necessity the subjective, our own consciousness. For this alone is that which is immediate; everything else, be it what it may, is first mediated by consciousness, and therefore dependent on it. It is thus rightly considered that the philosophy of the moderns starts with Descartes as its father (WWR 2,3).
\end{quote}

All this dissolves, however, with the contention that Kant's 'modern consciousness' stems not from his repositioning of the subject but from his redefinition of time. For in making time the form of inner sense, Kant subordinates the certainty of interiority to the productive forces of temporalization. Transcendental philosophy, as we will see, abandons the Cartesian line by conceiving of an inside that is passive in relation to the immanent exteriority of time. Dismantling the security of interiority by discovering an outside line that divides the inside from within, the \textit{Critique of Pure Reason} riddles interiority with the difference between the receptive nature of what exists in time and the synthesizing processes of the form of time itself. With Kant, then, the certainty of self consciousness dissolves into questions about the relation of time to itself. Critical thought thus not only differentiates itself from the tradition of modem philosophy which stems from Descartes, it fractures its very foundations.

As we have seen in the discussion of the 'Transcendental Aesthetic', the first critique divides the inside from the outside through the distinction that it makes between space and time. Space, for Kant, as the form of outer sense, constitutes the structure in which the objects of the external world are presented to the mind. The form of interiority, on the other hand, is time. Kant here defines the outside as that which exists in space and the inside as that which occurs in time. Yet, as we have already noted, since the exteriority of space is mediated by the inner determinations of the mind, everything that is in space is also in time. For Kant, then, all our experiences, whether internal or external, are inescapably conditioned by the order and relations of time.

From the perspective of transcendental philosophy, Cartesian doubt has an implicit spatial bias. Descartes seems far more concerned with the ways in which the evil demon may be deceiving us about the external world of space than he is with the deception that might be taking place in the internal realm of time.\footnote{Though Descartes claims that amongst the things which he doubts is the "time in which things endure" (Meditations 1: 8) this has little practical consequence. For Descartes' reconstruction of certainty occurs inside the passage of temporal succession (that is the days in which the \textit{Meditations} unfold). Thus, though Descartes claims he is questioning time, his philosophical method does little to challenge the interior structure of time.} This becomes startlingly apparent when one recognizes that the \textit{Meditations}, structured around a series of successive days, is governed by a strict temporality.\footnote{Deleuze claims that this structure of the Meditation makes it "the first text to introduce time into philosophical discourse". For unlike previous philosophical texts, with Descartes, the unfolding of temporal succession has positive implications for what can or cannot be said. In Descartes, says Deleuze, "there is a temporality which has unfolded which meant that he could not say in the second what he will say in the fifth" (TLK, 4). There is undoubtedly an irony here. For despite the fact that time operates in the Meditations as a positive structuring principle, Descartes nevertheless ignores the operations of time when securing his philosophical foundations.} Duped by the comfort of self consciousness, Descartes seems blind to the fact that his entire philosophical method unfolds inside a structure that is conditioned by time. Kant's break with Descartes must be located here. It is impossible for Kant to make the inside the "true point of support for philosophy" since the thought of the transcendental insists that the inside is produced by the alterity of time. 

This notion of time as the productive force of interiority depends, first of all, on shifting the line which. separates the inside from the outside. The border which Descartes locates at the break between the certainty of self consciousness and the doubt which haunts our knowledge of the outside world is radically altered by the first critique. With Kant the division between interiority and exteriority, as we have seen, ultimately corresponds to the distinction between the empirical realm of experience and the transcendental plane, which is to say that in Kant, the inside is defined as that which occurs in time, while the outside is left to the only thing which escapes this interiority, that is the abstract and productive forces of time itself. 
In focusing on this distinction---
between time and what is in time---Kant dramatically diminishes the importance of self consciousness. For consciousness of ourselves, to use Schopenhauer's term for the basis of Descartes' philosophy, is, for Kant, relegated to the level of the empirical. It is an experience which happens in time. Recognizing that "I am" is a verb conditioned by temporality,\footnote{As Deleuze writes the assertion 'I am a thing which thinks' serves---however implicitly---to bind the I think to a determinate entity in time "We cannot say with Descartes 'I think therefore I am. I am a thing which thinks'. If it is true that the I think is a determination, it implies in this respect an indeterminate existence (\textit{I am}). But nothing so far tells us under what form this existence is determined by the I think: it is determinable only in time, under the form of time, thus as the existence of a phenomenal, receptive or changing ego (KCP, viii).} Kant thus questions the very core of the Cartesian formula. The proposition 'I think' or 'I exist thinking,' writes Kant, "is an empirical proposition" (CPR, 381). To quote from the first critique: 

\begin{quote}
    Consciousness of self according to the determinations of our state of inner perception is merely empirical, and always changing. No fixed and abiding self can present itself in this flux of inner appearances. Such consciousness is usually named \textit{inner sense}, or \textit{empirical apperception} (CPR, 136).
\end{quote}

Therefore, what Descartes perceived as an asylum from doubt, becomes in Kant a constantly changing "flux of inner appearances" which constitutes, not the stability of the I think, but the superficiality of the empirical ego.

The ego, then, according to Kant, is in a passive or receptive relation to the structuring principles which condition it. As Deleuze writes, "the Ego itself is in time, and thus constantly changing: it is a passive or rather receptive Ego, which experiences changes in time" (ECC, 29). No longer transparent to itself, the subject becomes an entity continually affected by something which it cannot reach. For, the experience of interiority, like all inner determinations of the mind, has necessarily been structured by time as the form of inner sense. The primary position granted to the self as an object of awareness is thus subordinated in Kant to the synthesizing processes of time. Caught inside the web of temporality, the being which doubts cannot serve as a foundation since, according to critical thought, we only know the ego after it has already been given a sequence, an order and a rhythm by time. Critical philosophy is thus led to an extremely strange assertion. For, according to Kant, the 'spontaneous' activity of the synthetic \textit{a priori} which characterizes the subject of the Copernican revolution must be separated entirely from the knowledge of what goes on in our minds.

In the \textit{Critique of Pure Reason} Kant describes this curious consequence as the paradox of inner sense. Foreshadowed in the 'Transcendental Aesthetic', but only fully brought to light in the 'Deduction,' the paradox of inner sense, writes Kant, arises from the fact that "this sense represents to consciousness even our own selves only as we appear to ourselves not as we are in ourselves. For we intuit ourselves only as we are inwardly affected, and this would seem to be contradictory since we would then have to be in a passive relation [of active affection] to ourselves" (CPR, 166). The self of transcendental philosophy, is paradoxically split, fractured by a line that separates the consciousness of an empirical ego from an unconscious transcendental I. Moreover, the riddle continues, we are only conscious of ourselves after we have been worked over by that part of ourselves that we can never know. For one only becomes aware of the ego after it has affected itself from within. For Kant, then, inner sense is governed by a complicated circuit in which the self is 'auto-affected' by 'another'.

%TODO: Check for punctuation
This paradox which haunts Kant's notion of interiority is, as we have seen, a necessary consequence of the transcendental method. For, the fundamental conjunction of critical thought, appearance, conditions of appearance, demands that one look behind empirical reality to the abstract forces which condition it. According to Kant's Copernican revolution our experience of ourselves is not given, it is produced. Thus, for Kant, knowledge of the empirical ego is not an answer but a riddle. For the concrete experience of interiority is incapable of revealing the abstract transcendental conditions of its own production. Thus it is the very nature of transcendental production that ensures the shift away from the subject to time. For, the only thing that can produce the experience of being in time is the operation of time itself.

Thus far, the transcendental notion of time has been described as an abstract and empty form defined as nothing but a pure ordinal sequence. Taken on its own, however, this description is severely limited. For it is in danger of missing the fundamentally productive nature of the transcendental. Spontaneous and active, time at its most abstract is characterized by a continuous process of production. The form of temporality, for Kant, is, above all, not a static eternity. For as we will see, time, in transcendental philosophy, is made up of blind acts of synthesis continually at work producing the experience of a world that exists in time.

These productive capacities of the transcendental form of time are outlined by Kant in his discussion of the three fold synthesis set out in the first part of the 'Transcendental Deduction'. Before exploring these synthesis in detail it is important to stress, once again, that Kantian synthesis occurs in a realm independent of experience. The three synthesis outlined in the Deduction are the \textit{a priori} machines that underlie experience. As Deleuze writes, they do "not bear on diversity as it appears in space and time, but on the diversity of space and time themselves. Indeed, without them, space and time would not be represented"(KCP, 15).

Kant's discussion of the three fold synthesis, then, is an attempt to uncover the productive constituting forces of the form of time itself. "All our knowledge," writes Kant in the introduction to the section, "is finally subject to time, the formal condition of inner sense. In it they must all be ordered, connected and brought into relation" (CPR, 131). The synthetic processes of the deduction are meant to explain the means through which this occurs. Kant's goal is to explicate how the transcendental form of time functions to put the world in time. The focus, then, is not on a realm of production that occurs in time, but rather on the 'spontaneous' activity that accounts for the abstract possibility of the very experience of time.

After cautioning the reader of the difficulties which will inevitably beset an enterprise "never before attempted," Kant begins his discussion with what he calls 'the synthesis of apprehension in intuition'. "Directed immediately upon intuition" (CPR, 131) the 'synthesis of apprehension' is located in the receptive domain of sensibility. It transforms our intuitions into a grid that renders them capable of being perceived. For according to Kant, as Deleuze makes clear, there is not only a diversity that exists in time but a diversity of time itself. The very possibility of perception requires that this diversity be synthesized. For intuition to take place the unrepresentable multiplicity of time must be made representable. The synthesis of apprehension fulfills this requirement by combining the manifold of intuition into the units that constitute moments in time. It thus operates, as Deleuze writes to "produce the different parts of time" (KCP, 15).

The second synthesis, the synthesis of reproduction in the imagination, is the act through which the imagination ensures the \textit{a priori} rules which determine the "sequence or coexistence" of appearances. It is, according to Deleuze, "the means by which we reproduce the preceding parts as we arrive at the ones following" (KCP, 15). Without this second synthesis the order and association of representations would be completely random.

\begin{quote}
    If cinnabar were sometimes red, sometimes black, sometimes light, sometimes heavy, if a man changed sometimes into this and sometimes into that animal form, if the country on the longest day were sometimes covered with fruit, sometimes with ice and snow, my empirical imagination would never find opportunity when representing red to bring to mind heavy cinnabar (CPR, 132).
\end{quote}

The synthesis of reproduction is thus the transcendental ground which ensures the reproducibility of experience. It is the means through which past appearances are duplicated. It thus assures the coexistence and order of that which appears in time.

The third and final synthesis Kant calls the synthesis of recognition in a concept. By relating the synthesized manifold to the concepts of the understanding the synthesis of recognition allows one to recognize what has been intuited and reproduced by the other synthetic operations. According to Kant it is only through this final synthesis that one gains knowledge of the objects of experience. For "that which constitutes knowledge," writes Deleuze, "is not simply the act by which the manifold is synthesized, but the act by which the represented manifold is related to an object (recognition: this is a table, this is an apple, this is such and such an object)" (KCP, 15). In the \textit{Critique of Pure Reason} it is the synthesis of recognition which ensures the continuity of time. For in order to recognize any given object as identical in different times there must be a synthesis capable of guaranteeing identity through time.

Through this final synthesis, Kant deduces what is known as the synthetic unity of apperception, the 'pure, original, unchangeable consciousness' (CPR, 136) which constitutes the Kantian 'I think'. For, the ability to recognize the same object as identical in different times requires, according to Kant, a unity of consciousness, an unchanging identity, which ensures the temporal continuity necessary for recognition to take place (from the side of the transcendental subject). Transcendental apperception, then, is the necessary "condition which precedes all experience, and which makes experience possible" (CPR, 136). Defined simply as "numerically identical' it is the synthetic operation which explains the continuity of experience, guaranteeing that all my experience belongs to me.\footnote{As Deleuze writes, "My representations are mine in so far as they are linked in the unity of a consciousness in such a way that the 'I think' accompanies them" (KCP, 15).}

As we have seen, it is with the notion of the transcendental I that Kant most clearly reveals his radical departure from the modern philosophical line that has its roots in Descartes. The security of self consciousness which serves as the ground for Cartesian thought rests on the identity that exists between the knowledge of the I think and the temporally determined, conscious entity that I am. The Kantian 'I' on the other hand, does not correspond to any determinate entity.\footnote{According to Deleuze, this notion that the I think is not attached to a determinate entity is at the heart of Kant's objection to Descartes. With the Cartesian cogito, writes Deleuze, the I think is an act of instantaneous determination, which implies an undetermined existence (\textit{I am}) and determines this existence as that of a thinking substance (I am a \textit{thing that thinks}). But how can the determination apply to the undetermined if we cannot say under what form it is 'determinable'? (ECC, 29) With this objection Kant profoundly challenges the Cartesian formulation which equates the empty representation 'I think' to the determinate substance 'I am a thing which thinks'.} For, as has already been noted, the self of transcendental philosophy, is paradoxically split, fractured by a line that separates the consciousness of empirical experience from the unconscious realm of the synthetic \textit{a priori}. As Deleuze writes, "the I and the Ego are thus separated by the line of time which relates them to each other, but under the conditions of a fundamental difference" (KCP, viii). By defining interiority through the split between time as an abstract synthetic process and the concrete determination of that which exists in time, transcendental philosophy conceives of a self that is based on the absolute difference between the transcendental I and the empirical ego.

Critique demarcates this difference by a rigorous determination of the boundaries of knowledge. The Ego, as consciousness of inner sense, admits knowledge whereas the I, for Kant, can never be known (but is always presupposed in knowing). Existing in a realm outside all experience, transcendental apperception is never the object of direct awareness. Devoid of all logical characteristics, the I, writes Kant, is nothing other than a numerical identity. Empty of all content, stripped of its ties to substance or personality the I becomes an empty slot. Necessary to explain the very possibility of experience the 'Transcendental I' can be thought but never known. Transcendental apperception is philosophy's final answer to the oracle at Delphi. For Kant, 'know thyself' has become a rigorous impossibility, an illegitimate use of reason.

There is then, in Kant, a radical difference between what we are and what we know ourselves to be. For according to the 'paradox of inner sense' we are only conscious of ourselves after we have been worked over by that part of ourselves that we can never know. Kant thus posits a subject that is exterior in relation to itself.\footnote{This notion of a subject in an exterior relation to itself has obvious resonance with the psychoanalytic distinction between the ego and the id. Freud himself recognized that his notion of the unconscious borrowed much from transcendental thought. The psychoanalytic assumption of unconscious mental activity appears to us, he writes in his essay \textit{The Unconscious}, as an extension of the corrections undertaken by Kant" (U, 173).} An empty synthetic process that accompanies all your representations, the I in Kant is so named not because it exists inside you but because wherever you are it is. In Deleuze's reading of Kant, this finds its expression in Rimbaud's formula 'I is another'.

What this section has sought to show is that this 'other' is time. Deduced from the temporal continuity that underlies experience, transcendental apperception is determined solely as the agent of the transcendental form of time. For Kant, "I is an act which constantly carries out the synthesis of time" (KCP, viii). Thus, while the Ego marks the interiority of the subject, the I can only be defined as the exteriority of time. In introducing time into thought, Kant inserts an irreparable fissure into the core of the subject. To quote Deleuze: "the form of interiority means not only that time is internal to us but that our interiority constantly divides us from ourselves, splits us in two" (KCP, ix).

According to Heinrich Heine "The name of Immanuel Kant has the
might of an exorcism". "His thought," he writes "was a revolution and one not wanting in horrors". Even night wandering spirits, are stricken in terror at the sight of Kant's \textit{Critique of Pure Reason}" (RPG, 107). Deleuze agrees, Kant's "thinking machine is absolutely frightening" (KST, 1). This fright arises from the fact that Kant has transformed the nature of philosophical horror, replacing Cartesian fear with a sort of transcendental dread against which the certainty that guards us against the demon is powerless. For Kant there is no security in self consciousness, for it is nothing but a mask. By implanting uncertainty at the core of interiority, Kant riddles even the cogito with doubt. Like the replicants in the film \textit{Blade Runner}, transcendental philosophy thus torments the subject with the taunting refrain 'you are not what you think you are'.\footnote{In \textit{Tarrying with the Negative}, Žižek writes of \textit{Blade Runner} as a Kantian film: "In Blade Runner Deckard, after learning that Rachel is a replicant who (mis)perceives herself as human, asks in astonishment: "How can it not know what it is?" We can see, now, how more than two hundred years ago, Kant's philosophy outlined an answer to this enigma (TN, 15).} With Kant, then, "everything happens as if the "enemy" of thought was within". This enemy---the transcendental outside---does not exist in another time or another space. Its exteriority consists only of the fact that it knows nothing of borders. Thus unlike Descartes demon it can neither be locked out nor contained. One no longer need be afraid that the 'reality' of outer perceptions is a trick, for Kant's 'modern consciousness of time' plagued us with a much more intimate fear. Surrounding us in an inaccessible exteriority which infiltrates the inside at the very level of its production, the Kantian outside (or the syntheses of time) not only constitutes the experience of the external world, it generates the very experience of ourselves. For, as Deleuze writes, with Kant "it is not time that is interior to us; it is we who are interior to time" (ECC, 31).

\section{The Schematism: Time and Abstraction}

n his essay \textit{Theatrum Philosophicum}, Michel Foucault proposes that all philosophy can be defined according to its underlying antagonism towards the principle figure of classical thought. "What philosophy," he writes, "has not tried to overturn Platonism?\dots Are all philosophies individual species of the genus AntiPlatonic? Does each begin with a declaration of this fundamental rejection?" (TP, 166). These questions, when applied to Kant, will serve as the guiding thread in the final pages of this chapter. For to follow Kant in his exploration of transcendental time demands, as Foucault suggests, a return to critique's initial declaration, its fundamental rejection of transcendence.

The \textit{Critique of Pure Reason}, as has already been noted, overturns Platonism by discovering an abstract form of time that has been freed from its subordination to movement. It would be a mistake, however, to assume that this implies that Kant's argument with Plato can be restricted to any particular zone or dialogue.\footnote{Plato's most explicit discussion of time is found in the dialogue 'Timeaus', but as we will see the Kantian engagement with Plato has implications which extend far beyond this particular dialogue.} For, the notion of abstract time does much more than transform the ancient conception of temporality. As this section will make clear, abstract time subverts Plato at its core by revolutionising the notion of abstraction.

